---
title: "Assignment 1 - Language Development in ASD - part 2"
author: "Sofie L. H. Frandsen, Kristian Isen, Astrid Lang Ledager"
date: "July 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Language development in Autism Spectrum Disorder (ASD)

Background: Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail: i) relying on actual naturalistic language production, ii) over extended periods of time. We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

We then want to test the language trajectory of child and parent over time.

This Markdown is structured in the following way:

1. The exercises: read them carefully. Under each exercise you will have to write your answers, once you have written and run the code. This is the part that you have to directly send to the teachers.
2. An (optional) guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results under the exercise part.
3. In exercise 4 you will be asked to create the best possible model of language development in TD and ASD children, picking and choosing whatever additional variables you want from the dataset. Next time, the models produced by the different groups will compete against each other to see who can produce the best model, so choose carefully!

You will have to have a github repository for the code and send the answers to Celine and Riccardo without code (but a link to your github repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

N.B. The following lines are a summary of the questions to be answered, the step-by-step instructions and tips are below.

## Exercise 1) Preliminary Data Exploration

Describe the participant samples in the dataset (e.g. by diagnosis, age, etc.). Do you think the two groups are well balanced? If not, what do you think was the reason?

[HERE GOES YOUR ANSWER]
In order to investigate the balance between the two groups, we performed t-tests in the form of linear models. We found significant differences between the two groups in age, ADOS, and the mother's MLU. However, TD children are on average 12 months younger than the ASD children, and the mothers of TD children speak a little more to their children than mothers of ASD children.

The reason for the age difference in the two groups might be that ASD children develop in different ways (both from each other, but also from TD children) - some early, some later, and therefore it wouldn't make sense to use the same age groups for the two groups.

I don't know why TD mothers talk more to their children than ASD mothers, as there is no significant difference between how much the children talk in each group.

ADOS is the severity of ASD symptoms, so it makes sense that the groups are significantly different from each other, as the ASD children should have a higher ADOS score than TD children.

On all other measures, the two groups are well balanced. 


### Exercise 2) Children learning language: the effects of time and ASD
Describe linguistic development in TD and ASD children in terms of Mean Length of Utterance (MLU)?



### Exercise 3) Child directed speech as a moving target
Describe how parental use of language changes over time in terms of MLU. What do you think is going on?


### Exercise 4) Looking into "individual differences" (demographic, clinical or cognitive profiles)
The dataset contains some additional variables characterizing the kids’ cognitive and clinical profile: ADOS (autism severity), MSEL EL (Expressive Language, that is, verbal IQ, or linguistic skills at first visit as assessed by a psychologist using Mullen Scales of Early Learning), MSEL VR (Visual Reception, used as a proxy for non verbal IQ at first visit), Age, Gender, Ethnicity. Would it make sense to add any of them to your model of linguistic trajectories? Create the best possible model (the one that best explain the data, with MLU as outcome). Next time your model will be tested on new participants, and we will proclaim a winner. Describe your strategy to select the best models (how did you choose the variables to include?) and send the code to Riccardo and Celine.

### [OPTIONAL] Exercise 5) Comment on how the three linguistic variables measure linguistic performance (the so-called "construct validity" of the measures). Do they express the same variance?


### Structure of the code chunks

Basic stuff:
- Loading the libraries
- Setting the directory and loading the data
- Look at the data (which variables are there? Are they in the right format?) and describe the participants (by diagnosis)

We will try to answer three questions:

- Do children with ASD develop language differently from non-ASD children?
- Do parents speak differently to children with ASD than to non-ASD ones?
- Which variables should we use to best explain the child linguistic performance?
  
### Loading the relevant libraries

Load necessary libraries : what will you need?

- e.g. something to plot with
- e.g. mixed effects models

```{r Load Libraries}
library(pacman)
p_load(lmerTest, ggplot2, MuMIn)
```

### Define your working directory and load the data

- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data}

getwd()
locpath=getwd()
setwd(locpath)
data = read.csv("ass1_LangASD.csv")

```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Visit, Number of words used, Number of unique words used, length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r}
str(data)

#create subset
data1 = subset(data, data$VISIT == 1) 

#same as t-test (because diagnosis is either one or the other)
age = lm(Age ~ Diagnosis, data1)
summary(age)
#significant = groups are not well matched: the TD kids are on average 12 months younger

ados = lm(ADOS1 ~ Diagnosis, data1)
summary(ados)
#significant = groups are not well matched (ofc): the TD kids' ADOS score are on average 13 points less

vIQ = lm(verbalIQ ~ Diagnosis, data1)
summary(vIQ)
#insignificant = groups are well matched

nvIQ = lm(nonVerbalIQ ~ Diagnosis, data1)
summary(nvIQ)
#insignificant = groups are well matched

mlu = lm(CHI_MLU ~ Diagnosis, data1)
summary(mlu)
#insignificant = groups are well matched

pmlu = lm(MOT_MLU ~ Diagnosis, data1)
summary(pmlu)
#significant = groups are not well matched: parents of TD kids speak a little more to their children

tokens = lm(tokens_CHI ~ Diagnosis, data1)
summary(tokens)
#insignificant = groups are well matched

types = lm(types_CHI ~ Diagnosis, data1)
summary(types)
#insignificant = groups are well matched

#use glm with binomial family when having categorical variable (only two)
gender = glm(Gender ~ Diagnosis, family = binomial, data1)
summary(gender)
#insignificant = groups are well mathced

#have to invert it
ethn = glm(Diagnosis ~ Ethnicity, family = binomial, data1)
summary(ethn)
#insignificant = groups are well matched


length(data[data$Diagnosis == "ASD", 1])
length(data[data$Diagnosis == "TD", 1])
#there are more TD children than ASD children

```

[REPORT THE RESULTS]

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r}

boxplot(CHI_MLU ~ Diagnosis*VISIT, col = c("white", "lightgrey"), data)

ggplot(data, aes(data$Age, data$CHI_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Age (months)", y = "Child MLU", title = "Child MLU over time") 

ggplot(data, aes(data$VISIT, data$CHI_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Visit number", y = "Child MLU", title = "Child MLU over time")


mCHI_MLU = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1+VISIT|SUBJ1), data = data)
summary(mCHI_MLU)

mCHI_MLUin = lmer(CHI_MLU ~ 1 + VISIT * Diagnosis + (1+VISIT|SUBJ1), data = data)
summary(mCHI_MLUin)

```

How would you evaluate whether the model is a good model?
*There are many ways to assess how good a model is.* 


How much variance is explained by the models? *Look at marginal R squared. Take the one that explains the most.*
```{r}
r.squaredGLMM(mCHI_MLU)
r.squaredGLMM(mCHI_MLUin)

```

The interaction model explains the most variance (35%).

*Can also*
*Look at assumptions - explain both central and extreme data?*
*Conceptual - do we need these elements in our model?*


- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r}
boxplot(MOT_MLU ~ Diagnosis*VISIT, col = c("white", "lightgrey"), data)

ggplot(data, aes(data$Age, data$MOT_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Age (months)", y = "Mother MLU", title = "Mother MLU over time") 

ggplot(data, aes(data$VISIT, data$MOT_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Visit number", y = "Mother MLU", title = "Mother MLU over time")


mMOT_MLU = lmer(MOT_MLU ~ 1 + VISIT + Diagnosis + (1+VISIT|SUBJ1), data = data)
summary(mMOT_MLU)

mMOT_MLUin = lmer(MOT_MLU ~ 1 + VISIT * Diagnosis + (1+VISIT|SUBJ1), data = data)
summary(mMOT_MLUin)
```



### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Celine.

The dataset contains some additional variables characterizing the kids’ cognitive and clinical profile: 
ADOS (autism severity), 

MSEL EL (Expressive Language, that is, verbal IQ, or linguistic skills at first visit as assessed by a psychologist using Mullen Scales of Early Learning), 

MSEL VR (Visual Reception, used as a proxy for non verbal IQ at first visit), 

Age, 

Gender, 

Ethnicity. 

Would it make sense to add any of them to your model of linguistic trajectories? Create the best possible model


```{r}

mCHI_MLU = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1+VISIT|SUBJ1), data = data)

m1 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + verbalIQ + (1+VISIT|SUBJ1), data = data)
summary(m1)
r.squaredGLMM(m1)

m2 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + verbalIQ + ADOS1 +  (1+VISIT|SUBJ1), data = data)
summary(m2)
r.squaredGLMM(m2)

m3 = lmer(CHI_MLU ~ 1 + VISIT + verbalIQ + ADOS1 + (1+VISIT|SUBJ1), data = data)
summary(m3)
r.squaredGLMM(m3)

m4 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + types_CHI + verbalIQ +  (1+VISIT|SUBJ1), data = data)
summary(m4)
r.squaredGLMM(m4)


```

REPORT OF THE RESULTS

*average error: how wrong we expect the model to be on average*

What is a good model for training?
the more complex the model, the better it is to explain the data (mostly).
conceptually - Think about - do we actually learn something from predicting mlu from unique words? it doesn't really add something new


What is a good model for testing?
the simpler the better.
replication crisis: it is hard to know what explaining models are good at predicting.


shared variance + residuals of both variables. the residuals explain what is on top of the shared variance for each variable. 


cross-validation:
folding the data (pie). train the model on 3/4 of the pie, test it on the 1/4 of the pie, then repeat the process for each quarter of the pie.
how many depends on:
how much data you have - the omre the bigger the number of slices
leave-one-out - issue with your assumption about representation of the population in your data
the more folds, the longer it takes to run
tip: we're looking at kids, not datapoints.