---
title: "Assignment 1 - Language Development in ASD - part 3"
author: "Sofie L. H. Frandsen, Kristian Isen, Astrid Lang Ledager"
date: "August 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time, then learning how to cross-validate models and finally how to systematically compare models.

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/ra99bdvm6fzay3g/demo_test.csv?dl=0
* Utterance Length data: https://www.dropbox.com/s/uxtqqzl18nwxowq/LU_test.csv?dl=0
* Word data: https://www.dropbox.com/s/1ces4hv8kh0stov/token_test.csv?dl=0

```{r Load Libraries}
library(pacman)
p_load(lmerTest, modelr, stringr, tidyverse, caret, cvTools, Metrics, ggplot2)
```

### Define your working directory and load the data

- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data}

getwd()
locpath=getwd()
setwd(locpath)
data = read.csv("ass1_LangASD.csv")

```

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data () and on the test data. Report both of them. Compare them. Discuss why they are different.

[HERE GOES YOUR ANSWER]

*How did your models from last time perform? In this exercise you have to compare the results on the training data () and on the test data. Report both of them. Compare them. Discuss why they are different.*

RMSE of the training data is 0.306, while the RMSE of test data is 0.284. RMSE of the training is higher than RMSE of the test data, which means that the model is better at predicting new data than it is at explaining the data.   


- recreate the models you chose last time (just write the code again and apply it to Assignment2TrainData1.csv)
```{r}
model = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + types_CHI + verbalIQ +  (1+VISIT|SUBJ1), data = data)
```

- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the functions rmse() and predict() )
    *RMSE(model, data) dont use predict*
```{r}
modelr::rmse(model, data)
```

- create the test dataset (apply the code from assignment 1 part 1 to clean up the 3 test datasets)
load the data:
```{r}
data1 = read.csv("demo_test.csv")
data2 = read.csv("LU_test.csv")
data3 = read.csv("token_test.csv")
```

rename variables:
```{r}
data1 = plyr::rename(data1, c("Child.ID" = "SUBJ", "Visit" = "VISIT"))
```

make the same:
```{r}
data2$VISIT = str_extract(data2$VISIT, "\\d")
data3$VISIT = str_extract(data3$VISIT, "\\d")
```

```{r}
data1$SUBJ1 = gsub("\\.", "", data1$SUBJ)
data2$SUBJ1 = gsub("\\.", "", data2$SUBJ)
data3$SUBJ1 = gsub("\\.", "", data3$SUBJ)
```

select variables:
```{r}
d1 = select(data1, SUBJ1, VISIT, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw)
d2 = select(data2, SUBJ1, VISIT, MOT_MLU, MOT_LUstd, CHI_MLU, CHI_LUstd)
d3 = select(data3, SUBJ1, VISIT, types_MOT, types_CHI, tokens_MOT, tokens_CHI)

```

merge datasets:
```{r}
d1and2 = merge(d1, d2, by = c("SUBJ1", "VISIT"))
df6 = merge(d1and2, d3, by = c("SUBJ1", "VISIT"))
```

```{r}
dfV1 = subset(df6, df6$VISIT == 1) 
dfV1 = select(dfV1, SUBJ1, ADOS, MullenRaw, ExpressiveLangRaw)

dfV1 = plyr::rename(dfV1, c("ADOS" = "ADOS1", "MullenRaw" = "nonVerbalIQ", "ExpressiveLangRaw" = "verbalIQ"))

df6 = merge(df6, dfV1, by ="SUBJ1")
df6 = subset(df6, select = -c(ADOS, MullenRaw, ExpressiveLangRaw))

```

```{r}
str(df6)
df6$SUBJ1 = as.factor(df6$SUBJ1)
levels(df6$SUBJ1) = 1:length(levels(df6$SUBJ1))
```

```{r}
df6$VISIT = as.numeric(df6$VISIT)
```

```{r}

df6$Gender = as.factor(df6$Gender)
df6$Gender = plyr::revalue(df6$Gender, c("1" = "M"))

df6$Diagnosis = plyr::revalue(df6$Diagnosis, c("A" = "ASD", "B" = "TD"))

```

- test the performance of the models on the test data 
```{r}
testmodel = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + types_CHI + verbalIQ +  (1+VISIT|SUBJ1), data = df6)
```

```{r}
modelr::rmse(testmodel, df6)
```

- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())



### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

[HERE GOES YOUR ANSWER]

For the cross-validated basic model, the RMSE of the test data is 0.809, while the RMSE of the train data is 0.341, which tells us that the model is better at explaining the data than predicting the new data. 

For the cross-validated improved model, the RMSE of the test data is 0.477, while the RMSE of the train data is 0.309, which tells us that the model is better at explaining the data than predicting the new data, but that the improved model is better at predicting the new data than the basic model. 

- Create the basic model of ChildMLU as a function of Time and Diagnosis (don't forget the random effects!).

```{r}
basicmodel = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1+VISIT|SUBJ1), data = data)
```

- Make a cross-validated version of the model. (Tips: google the function "createFolds";  loop through each fold, train a model on the other folds and test it on the fold)

```{r}
folds = createFolds(unique(data$SUBJ1), k = 5)
rmse_train = NULL
rmse_test = NULL
n = 1

for(i in folds){
  train = subset(data, !(data$SUBJ1 %in% i))
  test = subset(data, data$SUBJ1 %in%i)

  newmodel = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1+VISIT|SUBJ1), data = train)
  
  newpred = predict(newmodel, newdata = test, allow.new.levels = TRUE)
  rmse_test[n] = Metrics::rmse(newpred, test$CHI_MLU)
  rmse_train[n] = Metrics::rmse(train$CHI_MLU, fitted(newmodel))
  
  n = n+1
}

mean(rmse_test)
mean(rmse_train)

```


- Report the results and comment on them. 

The rmse of the test data is 0.809, while the rmse of the train data is 0.341, which tells us that the model is better at explaining the data than predicting the new data. 

- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.
```{r}
goodmodel = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + types_CHI + verbalIQ +  (1+VISIT|SUBJ1), data = data)
```

```{r}
folds = createFolds(unique(data$SUBJ1), k = 5)
rmse_train = NULL
rmse_test = NULL
n = 1

for(i in folds){
  train = subset(data, !(data$SUBJ1 %in% i))
  test = subset(data, data$SUBJ1 %in%i)

  newmodel = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + types_CHI + verbalIQ + (1+VISIT|SUBJ1), data = train)
  
  newpred = predict(newmodel, newdata = test, allow.new.levels = TRUE)
  rmse_test[n] = Metrics::rmse(newpred, test$CHI_MLU)
  rmse_train[n] = Metrics::rmse(train$CHI_MLU, fitted(newmodel))
  
  n = n+1
}

mean(rmse_test)
mean(rmse_train)

```
This model is better at predicting new data than the basic model, as it has a lower rmse for the test data.

- Bonus Question 1: How would you go comparing the performance of the basic model and the cross-validated model on the testing set? 

Use rmse?
```{r}
modelr::rmse(basicmodel, data)
modelr::rmse(goodmodel, data)
```

- Bonus Question 2: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 3: compare the cross-validated predictive error against the actual predictive error on the test data

### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.
(Tip: recreate the equation of the model: Y=Intercept+BetaX1+BetaX2, etc; input the average of the TD group for each parameter in the model as X1, X2, etc.). 

[HERE GOES YOUR ANSWER]

*How does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.*

Bernie's MLU is higher than the average TD child's MLU for all visits.

*How does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations?*

Bernie's actual MLU at visit 6 is 3.45, while his predicted MLU at visit 6 is 3.39, so Bernie is a little above expectations at visit 6.

__

So: 
exclude the ASD children.

How does the average TD child perfrom.
Want six datapoints (one for each visit)

Does he fit in the predictive intervals?

Can look at summary of the model.
MLU = intercept estimate + 1*visit + 1*diagnosis + 1*verbalIQ

Extract Bernie (number 2) from test data set:
```{r}
Bernie = subset(df6, df6$SUBJ1 == 2)
```

Extract the TD children from the data set:
```{r}
TDchildren = subset(data, data$Diagnosis == "TD")
```

Find the mean verbalIQ:
```{r}
mean(TDchildren$verbalIQ)
```

Find the mean of types_CHI for each visit:
```{r}
mean(TDchildren$types_CHI[TDchildren$VISIT == "1"])
mean(TDchildren$types_CHI[TDchildren$VISIT == "2"])
mean(TDchildren$types_CHI[TDchildren$VISIT == "3"])
mean(TDchildren$types_CHI[TDchildren$VISIT == "4"])
mean(TDchildren$types_CHI[TDchildren$VISIT == "5"])
mean(TDchildren$types_CHI[TDchildren$VISIT == "6"])
```


```{r}
model1 = lmer(CHI_MLU ~ 1 + VISIT + types_CHI + verbalIQ + (1+VISIT|SUBJ1), data = TDchildren)
summary(model1)
```

```{r}
#MLU = 0.491 + (0.154*VISIT) + (0.007*types_CHI) + (0.018+verbalIQ)

#For visit 1, 2, etc:
MLU1 = 0.491 + (0.154*1) + (0.007*44.75)  + (0.018*20.151)
MLU2 = 0.491 + (0.154*2) + (0.007*94.281) + (0.018*20.151)
MLU3 = 0.491 + (0.154*3) + (0.007*133) + (0.018*20.151)
MLU4 = 0.491 + (0.154*4) + (0.007*161.063) + (0.018*20.151)
MLU5 = 0.491 + (0.154*5) + (0.007*175.355) + (0.018*20.151)
MLU6 = 0.491 + (0.154*6) + (0.007*174.607) + (0.018*20.151)

MLU1
MLU2
MLU3
MLU4
MLU5
MLU6

```

```{r}
BMLU1 = Bernie$CHI_MLU[Bernie$VISIT == "1"]
BMLU2 = Bernie$CHI_MLU[Bernie$VISIT == "2"]
BMLU3 = Bernie$CHI_MLU[Bernie$VISIT == "3"]
BMLU4 = Bernie$CHI_MLU[Bernie$VISIT == "4"]
BMLU5 = Bernie$CHI_MLU[Bernie$VISIT == "5"]
BMLU6 = Bernie$CHI_MLU[Bernie$VISIT == "6"]

BMLU1
BMLU2
BMLU3
BMLU4
BMLU5
BMLU6

```

Distance between Bernie's MLU and average TDs:
```{r}
MLU1-BMLU1
MLU2-BMLU2
MLU3-BMLU3
MLU4-BMLU4
MLU5-BMLU5
MLU6-BMLU6
```
Bernie's MLU is higher than the average TD child's MLU for all visits.

Plot:
```{r}
TD_meanMLU_est = data.frame("VISIT" = c(1, 2, 3, 4, 5, 6), "MLU_est" = c(1.32, 1.82, 2.25, 2.6, 2.85, 3))


mlu_TD = ggplot(data = TD_meanMLU_est, aes(x = VISIT, y = MLU_est)) + geom_bar(stat = "identity", width = 0.15, fill = "grey", colour = "black")
mlu_TD_andBernie = mlu_TD + geom_line(data = Bernie, aes(x = as.numeric(VISIT), y = CHI_MLU), colour = "red") + theme_minimal()
mlu_TD_andBernie
```

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r}
predBMLU = predict(model1, Bernie, allow.new.levels = TRUE)
Bernie$CHI_MLU[6]
predBMLU[6]
```

Bernie's actual MLU at visit 6 is 3.45, while his predicted MLU at visit 6 is 3.39, so Bernie is a little above expectations at visit 6.


### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!


